---
layout: post
title: "Mapping waxwings annual migration without Twitter"
comments: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE, 
                      cache = FALSE)
```

Recently a reader left a comment on this blog mentioning [his cool blog post](https://robbriers.github.io/post/2017-03-14-mapping-the-spread-of-a-migratory-bird-using-twitter/) in which he mapped the spread of a migratory bird using Twitter. His data source was [the Waxwings UK account](https://twitter.com/WaxwingsUK) which reports sightings of Waxwings in the UK. I decided to try reproducing and extending his work using [the rOpenSci `spocc` package](https://github.com/ropensci/spocc) that interfaces different sources of species occurrence data.

<!--more-->

# Getting the occurrence data

As mentioned above I got the data via `spocc`. I read the README of the Github repo and learnt that it's called `spocc` like *sp*ecies *occ*urrence data. So now I should never forget how many "c" there are in the word _occurrence_. Now please send help for my remembering it has two "r".

The `spocc` package interacts with so many data sources that I felt a bit overwhelmed. I guess ecologists are not often blessed with so much data. Note that I don't limit my query to the UK. Indeed, since seeing [this map](https://en.wikipedia.org/wiki/Bohemian_waxwing#/media/File:Bombycillagarrulusmap2.png) my interest for the Bohemian waxwing became global. I even decided to get data for the two other species of waxwing, although it only worked for the Bohemian and Cedar waxwing in the end.

I decided to use only data from [GBIF](http://www.gbif.org/). You'll find more information about GBIF for instance in this [blog post from rOpenSci blog](https://ropensci.org/blog/2014/03/17/spocc/) Since I wanted to get data for different years and all 3 species of waxwings, which meant a lot of data so for getting it I sliced the Earth (I like how evil this sentence makes me sound, ah!). Note that I used overlapping slices because otherwise I didn't get data at the limits between slices. I guess I could have made my slices slightly less overlapping though.

```r
library("spocc")
library("dplyr")
library("purrr")
library("scrubr")
get_slice <- function(longitude, year){
  print(paste(year, longitude))
  gbifopts <- list(limit = 200000,
                   year = year)
  waxwings <- occ(query = "Bombycilla",
                  from = c('gbif'), 
                  gbifopts = gbifopts,
                  geometry = c(longitude - 0.5, - 90, 
                               longitude + 0.5, 90))
  waxwings <- occ2df(waxwings)
}
longitudes <- seq( -180, 179, by = 0.5)
years <- rep(2011:2016, length(lo ngitudes))
longitudes <- rep(longitudes,  6)
waxwings <- map2(longitudes, years, get_slice)
for (i in 1:length(waxwings)){
  waxwings[[i]]$latitude <- as.numeric(waxwings[[i]]$latitude)
  waxwings[[i]]$longitude <- as.numeric(waxwings[[i]]$longitude)
}

waxwings <- bind_rows(waxwings)
waxwings <- unique(waxwings)

waxwings <- filter(waxwings, !is.na(name))
readr::write_csv(waxwings, path = "uncleaned_waxwings.csv")


```

# Cleaning the occurrence data

Now because my MSc of ecology is far behind me (but still close to my heart!) I have no idea how to assess the quality of species occurrence data. And even if I did I would have been delighted by the discovery of another rOpenSci package, [`scrubr`](https://github.com/ropensci/scrubr), whose aim is to clean species occurrence records. Because I had so much data, I cleaned each day separately, otherwise it was just too long and hard on my poor computer. Don't start thinking `scrubr` is slow, it isn't and it's getting faster by the minute [thanks to work by its maintainer](https://github.com/ropensci/scrubr/issues/26).

```r
cleanup <- function(df){
  print(df$date[1])
  df <- df %>%
    coord_impossible() %>%
    coord_incomplete() %>%
    coord_unlikely() 

    if(nrow(df) > 1){
    
      df <- dedup(df)
    }
  df <- df %>%
    date_standardize("%Y-%m-%d") %>%
    date_missing()
    return(df)
} 
waxwings <- readr::read_csv("uncleaned_waxwings.csv")


waxwings <- split(waxwings, 
                  waxwings$date)

waxwings <- lapply(waxwings, cleanup)

waxwings <- bind_rows(waxwings)
waxwings <- unique(waxwings)
waxwings <- filter(waxwings, name != "Bombycilla japonica",
                   longitude < 50)
readr::write_csv(waxwings, path = "waxwings.csv")
```

```{r, eval = TRUE, echo = FALSE}
waxwings <- readr::read_csv("waxwings.csv")
waxwings <- filter(waxwings, name != "Bombycilla japonica",
                   longitude < 50)

```

I removed the records with impossible, incomplete or unlikely coordinates (unlikely being e.g. a political centroid, impossible coordinates with a too high longitude), I also removed duplicate records and records without a data. This was so easy! I'd like the `scrubr` package to come clean my flat, as a team with the [`janitor` package](https://github.com/sfirke/janitor). Now, in real life, I'd probably be even stricter with data cleaning but for the scope of this blog post, using that package without any additionnal check was enough.

I also removed occurrences of the Japanese waxwing because they were too few of them, and occurrences with a longitude higher than 50 because it seemed weird to have non Japanese waxwings in Asia. After all this glorious data cleaning, I had `r nrow(waxwings)` records. 

# Exploring the data

## Number of occurrences by species over time

Seasonality (fewer people where they breed)

## Day-of-the-week effects

More occurrence reports in the week-ends. Mention the bizday package I could have used for finer characterization in particular in the US (because taking into account different holidays in different countries, huh.)

# Mapping the migrations!

I first decided to plot the occurrences themselves on maps, by month, and to make a gif out of it. Pretty basic choice. I then had to choose the colour and shape of the points used to represent the birds. Shape? Bird emojis of course! Regarding the colour, I was quite glad when someone I follow on Twitter posted about the birdcolourbot account created by [David Lawrence Miller](https://twitter.com/millerdl/status/846713149323726848), because I looked and found [this tweet](https://twitter.com/birdcolourbot/status/847706299936587776) with colours for the bohemian waxwing, Bombycilla garrulus, the one present in both America and Europe. For the cedar waxwing I had to create a palette myself, which I'd never thought of doing if I hadn't seen the birdcolourbot account. I used [a random pic](https://www.allaboutbirds.org/guide/Cedar_Waxwing/id) from the internet. In both cases, I uploaded images [on this website](http://imagecolourpicker.com/es) to get colour codes. A bit of copy-paste work but much better than having to mix paint material for real.

```{r}
bohemian_palette <- c("#1A1A1A", "#878787",
                    "#B15929", "#E21A1C",
                    "#FEFF99")

cedar_palette <- c("#050608", "#5D5C7A","#AF5F2A",
                   "#8E2F49",
                   "#F3DD31")

```

I think both species look quite similar so I didn't expect the palettes to be reallt different. I decided the colours would all have the same probability, instead of weighing them according them to their presence in the usual patterns of each species.

```{r}
set.seed(1)
library("dplyr")
bohemian <- filter(waxwings,
                   name == "Bombycilla garrulus")
bohemian <- mutate(bohemian,
                   colour = sample(bohemian_palette, size = nrow(bohemian), replace = TRUE))

cedar <- filter(waxwings,
                   name == "Bombycilla cedrorum")
cedar <- mutate(cedar,
                   colour = sample(cedar_palette, size = nrow(cedar), replace = TRUE))

```

Animated maps

replace ugly red points by bird emojis

maybe try spatial smoothing by month... or mention it as a possible extension.

maybe use a better projection.

# Concluding

I was impressed by rOpenSci tools for getting and cleaning occurrence data. 